animbook: Visualizing Changes in Performance Measures and Demographic Affiliations using Animation

### 2025-03-20 response_to_review.qmd

---
title: "Response to Review"
subtitle: "animbook: Visualizing Changes in Performance Measures and Demographic Affiliations using Animation"
author: "Krisanat Anukarnsakulchularp, Dianne Cook"
format: pdf
---

We thank the editor and reviewers for reading and offering improvements for our article. These are the changes we have made in response. *The reviewer comment is in italics*, and our response is regular text.

## Reviewer 2

### *General*

*The article explores the use of animated ggplot2 plots with the animbook package. This package was inspired by an animated graphic from the New York Times and has the benefit of helping to visualize trends over time. The package takes a three stage approach, using a data cleaning step, then graphing step, and finally an animation step. The end result is an animated graphic that shows the relationship of categorical data over time.*

Response:

### *Detailed comments*

*Overall, the authors did a good job at guiding the narrative in a logical fashion. One major recurring theme that I noticed was that paragraphs tended to be short and had uneven transitions. My comments are below and follow the progression of the paper.*

- *Page 1, second paragraph: this paragraph is really short and would benefit by describing the animation in detail.*

Response:

- *Page 2, paragraph starting with Hvitfeldt: it might be helpful to include parts of the example from Hvitfeldt, or at least describe what parts of the example inspired animbook*

Response:

- *Page 2, paragraph starting with "In the data structure": I suggest a little more pretext for the example of company name and where the company is registered. Or state the purpose of ID and grouping before giving a cohesive example.*

Response:

- *Page 3, paragraph starting with "One of the frequently": I was initially confused about how NA were handled, but it was clearer later in the paper that NA became its own category.*

Response:

- *Page 3, paragraph starting with "A principle important": switch to "An important principle"*

Response:

- *Page 3, paragraph starting with "There are multiple": avoid single sentence paragraphs. This could be included at the end of the previous paragraph to make a cleaner transition.*

Response:

- *Page 4, paragraphs 2 and 3: these paragraphs could be combined*

Response:

- *Page 5, paragraph 2: I don't think that redundancy needs to be capitalized here*

Response:

- *Page 5, last paragraph of section 5: "As illustrated in Figure 5." is left as its own sentence, but should be included with the surrounding sentence.*

Response:

- *Page 5, last sentence: change "Github" to "GitHub". Also, it would be helpful to include the GitHub link here as well.*

Response:

- *Page 6, Data pre-processing section: "ranking the sales" should be replaced with "ranking the observations".*

Response:

- *Page 6: the sentence with "When the observation", can be improved for clarity ("When an observation is moved up by a quantile, another one is bound to ...")*

Response:

- *Page 6: Place tibble into table with a caption*

Response:

- *Page 7, first paragraph: replace "from the raw into" to "from the raw format into"*

Response:

- *Page 7, 3rd paragraph: include this information in another paragraph to help transitions.*

Response:

- *Page 7, 3rd and 4th paragraphs: the time_dependent argument only appears in the _plot() functions, so it should be mentioned in the later section.*

Response:

- *Page 8, table 2: "argumenst" should be "arguments"*

Response:

- *Page 8, paragraph 2/3: the proportional_shade() function can be discussed solely in one paragraph for better transition.*

Response:

- *Page 8, paragraph 4: should "wallaby's plot" be "wallaby plot"?*

Response:

- *Page 8, last paragraph: state that the data is included in the animbook package.*

Response:

- *Page 9: remove comma after "Figure 7"*

Response:

- *Page 10, applications: include sources at the first mention of each example.*

Response:

- *Page 13: use consistent naming for the plot objects (kan_p vs. p)*

Response:

- *Page 15, paragraph starting with "In Figure 11": There is a sentence prompt that should be filled in "(explain the difference in proportion ...)"*

Response:

- *Page 16, the two paragraphs before discussion: the second paragraph leads to explanation from the previous paragraph, so these ideas should stay together in one paragraph.*

Response:

- *Page 18: There is a reference starting with "---------" that should be filled in.*

Response:

- *An example with the funnel_web_plot() would be a nice addition in the applications section.*

Response:

## Code

*Usability:*

- *Page 6: it states that users can specify breaks on a 0 to 1 scale. Is there a reason why users can't provide breaks on the raw data scale and have the function transforms the values to fit the 0 and 1 scale?*

Response:

- *Page 8: could the last_plot() function be used by default so that users don't have to save the plot into a ggplot2 object?*

Response:

- *In the wallaby plot, the applicable points move from all categories to a single category (or vice versa). Is there a way to choose which single category gets used?*

*Code:*

- *When using custom breaks, the error displays "The breaks vector must have the same number of groups as ncat argument", but it is stated in the article that the length of breaks need to equal the number of categories plus 1.*

Response:

- *When running the code on an M1 MacBook Pro, some of the shaded sections tear apart and look like white circles. This happens particularly with the wallaby plots.*

Response:

### *Other comments:*

- *Are there plans to include additional visuals to help keep track of the points as they disappear? In the NYT article, the points are summed and proportions are calculated as the points reach the right side of the plot and I think this would be a nice addition in the future.*

Response:

- *When there are many data points, it becomes difficult track the general movement of observations. Would there be any options to trace the paths to show where the points have moved (like a line with low alpha)?*

Response:

- *Overall, I really like this package!*

Response: Thank you!

## Reviewer 3

### *General*

*The paper presents a graphical tool for visualizing changes in some measurements over time. It also allows for multiple groups to appear in the Figure. At first glance, this sounds like a useful tool. Me personally, I was not amazed by this tool simply because of the lack of simplicity and straightforward interpretation. On the positive side, the paper is very professionally written, but it is not that easy to follow.*

Response:

### Detailed comments

*1. Figure 1 for instance. What is the interpretation of this figure? I struggled to understand and did not manage to do so.*

Response:

*2. Figures 2 and 4. Since the data entered in R are numerical or categorical values, what is the use of these figures? Why should the reader be informed of the raw, handwritten data?*

Response:

*3. I have spotted a few grammatical errors:*

*a) First paragraph after Figure 1. “This data…” > “These data…”. The word data is plural, the singular form is datum.*

Response:

*b) Beginning of Section 5. Please delete the letters “R. E.” from the reference of Mayer and Moreno (2002).*

Response:

*c) Page 13. First paragraph, 3rd and 4rth lines probably have a typo (“…and japan”).*

Response:

*4. The description you give for Figure 10 helps the reader understand the meaning of this plot. But, in the example in the R package with the aeles dataset, things are not so obvious. There are many observations that mask the visibility of the points.Also, what do the X and Y mean?*

Response:

*5. Why do the authors write “I acknowledge…” and not “We…” in the acknowledgements?*

Response:

### *Code*

*I did not find the commands in the package very user-friendly. For instance, to produce a plot I must go through many functions, prior to typing the plot to appear. The example in the aeles dataset is such a case. I need to type 4 commands in order to see a plot, which I repeat, cannot understand, nor interpret easily. I would prefer a function that does all this preprocessing and plots the data. Legends are missing, and in my opinion, the help files are not very helpful for someone who wants to use this package easily, to visualize their data.*

Response:

### 2025-04-29 response_to_review.qmd

### *Detailed comments*

*Overall, the authors did a good job at guiding the narrative in a logical fashion. One major recurring theme that I noticed was that paragraphs tended to be short and had uneven transitions. My comments are below and follow the progression of the paper.*

- *Page 1, second paragraph: this paragraph is really short and would benefit by describing the animation in detail.*

Response: We have added a short description of the New York Times animation.

@@

- *Page 2, paragraph starting with Hvitfeldt: it might be helpful to include parts of the example from Hvitfeldt, or at least describe what parts of the example inspired animbook*

Response: We have emphasized that his example code that reproduces the New York Times visualization inspires this package.

@@

- *Page 2, paragraph starting with "In the data structure": I suggest a little more pretext for the example of company name and where the company is registered. Or state the purpose of ID and grouping before giving a cohesive example.*

Response: ??? I have added US and Japan as an example, not sure what else can be added.

@@

- *Page 8, table 2: "argumenst" should be "arguments"*

Response: Fixed

- *Page 8, paragraph 2/3: the proportional_shade() function can be discussed solely in one paragraph for better transition.*

Response: ??? I feel like the function should be mentioned first in the second paragraph. The third paragraph is mainly about the function algorithm.

@@

*1. Figure 1 for instance. What is the interpretation of this figure? I struggled to understand and did not manage to do so.*

Response: Describe what do we learn.

*2. Figures 2 and 4. Since the data entered in R are numerical or categorical values, what is the use of these figures? Why should the reader be informed of the raw, handwritten data?*

Response: Not change, this image is important base on discussion with user. (see how other use the package). (be more forceful)

@@

*a) First paragraph after Figure 1. “This data…” > “These data…”. The word data is plural, the singular form is datum.*

Response: Just change.

*b) Beginning of Section 5. Please delete the letters “R. E.” from the reference of Mayer and Moreno (2002).*

Response: All referencing are automated and the style is consistent with the R journal style so we have no control over this.

### 2025-04-29 animbook-journal.Rmd
Reviwer 2 page 13

---
title: "animbook: Visualizing Changes in Performance Measures and Demographic Affiliations using Animation"
abstract: >
  Performance is often reported in quantiles or ranks and often repeatedly provided over time. Between elections, voters often switch party preferences, which produces categorical values over time. Inspired by the animation in the New York Times article "Extensive data show punishing reach of racism for black boys," we provide new visualization functions to generate animations that display this type of data in an R package called animbook. Functions to pre-process the data and prepare it for the animation are provided. The methods will be suitable generally for data where categorical variables are measured over time, and provide a way to engagingly view changes.
author:  
  - name: Krisanat Anukarnsakulchularp
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    - Melbourne, Australia
    orcid: 0009-0008-5638-7124
    email:  kanu0003@student.monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address: 
      - Department of Econometrics and Business Statistics
      - Melbourne, Australia
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: animbook-journal.bib
---

@@

  The New York Times provided a possible solution in the article titled "Extensive data show punishing reach of racism for black boys" (@the_new_york_time) to tell the story of how racism appears to inhibit socioeconomic change. This animation is the motivation for the new visualization presented here to be applied generally.

@@

  @Hvitfeldt_2018 provides R (@r) code to generate an animation similar to that of the New York Times. Taking inspiration from this example, \CRANpkg{animbook} expands upon the visualization in numerous ways and packages the code so that it can be used for a variety of similar animations for studying flow and temporal change data.

@@

In the data structure, there are requirements that must be followed for reproducing the animation. First, the data set needs to be in a `tidy data` format (@tidy-data). The data then must have at least ID and time variables, in addition to the measured variables, which would usually be numeric but can also be categorical as well. The ID variable indicates the individual, which is followed over time, such as the company name. There may also be a grouping variable, such as the country where the company is registered.

@@

  One of the frequently encountered values in the dataset is NA, commonly known as a missing value. In the context of accounting data, NA might signify that a company is either not yet listed or has been de-listed. In scenarios like voter data, NA may indicate that individuals did not provide the necessary information. Disregarding NA values could result in the loss of valuable information. Therefore, instead of completely removing NA entries, this information is categorized into its quantiles, ensuring that the missing data are factored into the visualization.

@@

A principle important for designing a useful animation is called persistence of vision (@animation-mechanic). When an image disappears, the brain will retain the previous images for a brief period of time. It is this slight period of retention that allows humans to separate sequential images. If this is seen in quick succession, it will appear as if the objects are in motion. This is illustrated in Figure \@ref(fig:animation-diagram).

  There are multiple ways to create an animation in the R environment (@r), including the packages \CRANpkg{gganimate} (@gganimate) and \CRANpkg{plotly} (@plotly).

@@

  In the context of data, as shown in Figure \@ref(fig:animated-diagram), observations are positioned at specific points in time. The further the distance between these points, the less smooth the animation becomes. This issue can be eliminated by interpolating additional points in between the observations. In the \CRANpkg{gganimate} package, the interpolation is achieved using the \CRANpkg{tweenr} package (@tweenr), while in \CRANpkg{plotly}, it utilizes d3.interpolate (@d3js).

  Figure \@ref(fig:animated-diagram) demonstrates how the frame variables are applied in an animated plot. The frame variable within the animated data structure allows the animation function to determine the position of observations on the plot at any given frame.

@@

  The paths of observations are also interpolated to create small changes in the position of points to produce the appearance of motion. The interpolation functions in \CRANpkg{gganimate} and \CRANpkg{plotly} are inadequate because they only generate a linear path. Our animation needs a non-linear path to show the flow. Sigmoid curves are commonly used in Sankey diagrams (@Hvitfeldt_2018). However, @Shaffer_2019 argues that a sine curve more accurately represents proportion throughout the path. As illustrated in Figure \@ref(fig:proportional-shade). The sigmoid accurately represents the proportion at the beginning and end, but as it curves, the shape gets narrower, leading to a less accurate proportion representation than the sine curve. New interpolation functions are needed to compute both the sigmoid and sine paths.

@@

\CRANpkg{animbook} is available through CRAN. The development version of this package can be found in the Github repositories.

@@

From Figure \@ref(fig:data-diagram), there is a need to map numerical value to a category. One way to handle this is by ranking the sales and grouping the rankings into quantiles. In some cases, this may not be the best option. When the observation is moved up by quantile, one is bound to move down, making the amount of observation in the quantile remain the same. This issue can be resolved by using an alternative method, which is grouping values based on their absolute values. Users may also be interested in grouping the data based on different groups, for example, ranking within a specific country. This generalization leads to four different scaling methods for the numerical data.

@@

```{r}
original |> 
  left_join(rank_data, by = c("id", "time", "gp")) |> 
  left_join(rank_group_data, by = c("id", "time", "gp")) |> 
  left_join(absolute_data, by = c("id", "time", "gp")) |> 
  left_join(absolute_group_data, by = c("id", "time", "gp")) |> 
  reframe(id, time, gp, values, rank, rank_group, absolute, absolute_group)
```

@@

  These are only the initial steps in formatting the data into a category. Now that there is a method to transform the data from the raw into a categorized format, the next step is to modify it into an animated data structure. It is carried out by assigning a frame to each individual observation, ensuring that each ID does not contain repeated frame values. It lets the \CRANpkg{gganimate} or \CRANpkg{plotly} perceive where the observation would be on the plot at a given frame, as seen in Figure \@ref(fig:animated-diagram).

@@

```{r tbl-cat-la, eval=knitr::is_latex_output()}
argument_num_tab |> 
  kableExtra::kable(caption = "The argumenst for the anim\\_prep\\_cat function.", format = "latex") |> 
  kableExtra::column_spec(2, width = "30em")
```

  The arguments for the `anim_prep()` and `anim_prep_cat()` functions are listed in Tables `r knitr::asis_output(ifelse(knitr::is_latex_output(), '\\ref{tab:tbl-num-la}','\\@ref(tab:tbl-num)'))` and `r knitr::asis_output(ifelse(knitr::is_latex_output(), '\\ref{tab:tbl-cat-la}','\\@ref(tab:tbl-cat)'))` respectively. The `data`, `id`, `values`, and `time` arguments are required for both functions.

  The `ncat`, `breaks`, `group_scaling`, and `scaling` arguments are for the user to customize the data scaling calculation in the `anim_prep()` function. In the `anim_prep_cat` function, the users can customize the order of the category to be shown on the y-axis using the `order` argument.

  Then, for further customization regarding how the final visualization looks. The `group`, `label`, and `time_dependent` arguments can be adjusted.

@@

The `cat_change` data in the package is used to illustrate how to use this package. It has five categories, A to E, over two time points. Since the data is already in the categorized data structure, the `anim_prep_cat()` function is used. The output from this function is a class called `categorized`.

@@

Figure \@ref(fig:top-many), shows the flexibility to use various options, such as `subset` or `relation`, to customize the plot appearance. Here `subset = "top"` is used to specify category A for the final time point, and `relation = "many_one"` is used to specify that all categories flowing into A at time point 2 are shown.

>>>>

---
title: "animbook: Visualizing Changes in Performance Measures and Demographic Affiliations using Animation"
abstract: >
  Performance is often reported in quantiles or ranks and often repeatedly provided over time. Between elections, voters often switch party preferences, which produces categorical values over time. Inspired by the animation in the New York Times article "Extensive data show punishing reach of racism for black boys," we provide new visualization functions to generate animations that display this type of data in an R package called animbook. Functions to pre-process the data and prepare it for the animation are provided. The methods will be suitable generally for data where categorical variables are measured over time, and provide a way to engagingly view changes.
author:  
  - name: Krisanat Anukarnsakulchularp
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    - Melbourne, Australia
    orcid: 0009-0008-5638-7124
    email:  kanu0003@student.monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address: 
      - Department of Econometrics and Business Statistics
      - Melbourne, Australia
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
type: package
output: 
  rjtools::rjournal_pdf_article:
    toc: no
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: animbook-journal.bib
---

@@

  The New York Times provided a possible solution in the article titled "Extensive data show punishing reach of racism for black boys" (@the_new_york_time) to tell the story of how racism appears to inhibit socioeconomic change represent using the animation showing an individual transition from one category to another. This animation is the motivation for the new visualization presented here to be applied generally.

@@

  @Hvitfeldt_2018 provides R (@r) code to reproduce an animation similar to that of the New York Times. Taking inspiration from this coding example, \CRANpkg{animbook} expands upon the visualization in numerous ways and packages the code so that it can be used for a variety of similar animations for studying flow and temporal change data.

@@

In the data structure, there are requirements that must be followed for reproducing the animation. First, the dataset needs to be in a `tidy data` format (@tidy-data). The data then must have at least ID and time variables, in addition to the measured variables, which would usually be numeric but can also be categorical as well. The ID variable indicates the individual, which is followed over time, such as the company name. There may also be a grouping variable, such as the country where the company is registered (United States and Japan).

@@

  One of the frequently encountered values in the dataset is NA, commonly known as a missing value. In the context of accounting data, NA might signify that a company is either not yet listed or has been de-listed. In scenarios like voter data, NA may indicate that individuals did not provide the necessary information. Disregarding NA values could result in the loss of valuable information. Therefore, instead of completely removing NA entries, it will get converted into its own quantiles, ensuring that the missing data are factored into the visualization.

@@

An principle important for designing a useful animation is called persistence of vision (@animation-mechanic). When an image disappears, the brain will retain the previous images for a brief period of time. It is this slight period of retention that allows humans to separate sequential images. If this is seen in quick succession, it will appear as if the objects are in motion. This is illustrated in Figure \@ref(fig:animation-diagram). There are multiple ways to create an animation in the R environment (@r), including the packages \CRANpkg{gganimate} (@gganimate) and \CRANpkg{plotly} (@plotly).

@@

  Figure \@ref(fig:animated-diagram) demonstrates how the frame variables are applied in an animated plot. The frame variable within the animated data structure allows the animation function to determine the position of observations on the plot at any given frame. These observations are positioned at specific points in time. The further the distance between these points, the less smooth the animation becomes. This issue can be eliminated by interpolating additional points in between the observations. In the \CRANpkg{gganimate} package, the interpolation is achieved using the \CRANpkg{tweenr} package (@tweenr), while in \CRANpkg{plotly}, it utilizes d3.interpolate (@d3js).

@@

  The paths of observations are also interpolated to create small changes in the position of points to produce the appearance of motion. The interpolation functions in \CRANpkg{gganimate} and \CRANpkg{plotly} are inadequate because they only generate a linear path. Our animation needs a non-linear path to show the flow. Sigmoid curves are commonly used in Sankey diagrams (@Hvitfeldt_2018). However, @Shaffer_2019 argues that a sine curve more accurately represents proportion throughout the path. As illustrated in Figure \@ref(fig:proportional-shade), the sigmoid accurately represents the proportion at the beginning and end, but as it curves, the shape gets narrower, leading to a less accurate proportion representation than the sine curve. New interpolation functions are needed to compute both the sigmoid and sine paths.

@@

\CRANpkg{animbook} is available through CRAN. The development version of this package can be found in the GitHub repositories at [https://github.com/KrisanatA/animbook](https://github.com/KrisanatA/animbook).

@@

From Figure \@ref(fig:data-diagram), there is a need to map numerical value to a category. One way to handle this is by ranking the observation and grouping the rankings into quantiles. In some cases, this may not be the best option because when an observation is moved up by a quantile, another one is bound to move down, making the amount of observation in the quantile remain the same. This issue can be resolved by using an alternative method, which is grouping values based on their absolute values. Users may also be interested in grouping the data based on different groups, for example, ranking within a specific country. This generalization leads to four different scaling methods for the numerical data.

@@

```{r}
scaling_table <- original |> 
  left_join(rank_data, by = c("id", "time", "gp")) |> 
  left_join(rank_group_data, by = c("id", "time", "gp")) |> 
  left_join(absolute_data, by = c("id", "time", "gp")) |> 
  left_join(absolute_group_data, by = c("id", "time", "gp")) |> 
  reframe(id, time, gp, values, rank, rank_group, absolute, absolute_group)
```

```{r tbl-scaling, eval=knitr::is_html_output()}
scaling_table |> 
  knitr::kable(caption = "A summary of different scaling methods applied to the dataset.")
```

```{r tbl-scaling-la, eval=knitr::is_latex_output()}
scaling_table |> 
  knitr::kable(caption = "A summary of different scaling methods applied to the dataset.") |> 
  kableExtra::column_spec(8, width = "30em")
```

@@

  These are only the initial steps in formatting the data into a category. Now that there is a method to transform the data from the raw format into a categorized format, the next step is to modify it into an animated data structure. It is carried out by assigning a frame to each individual observation, ensuring that each ID does not contain repeated frame values. It lets the \CRANpkg{gganimate} or \CRANpkg{plotly} perceive where the observation would be on the plot at a given frame, as seen in Figure \@ref(fig:animated-diagram).

@@

```{r tbl-cat-la, eval=knitr::is_latex_output()}
argument_num_tab |> 
  kableExtra::kable(caption = "The arguments for the anim\\_prep\\_cat function.", format = "latex") |> 
  kableExtra::column_spec(2, width = "30em")
```

  The arguments for the `anim_prep()` and `anim_prep_cat()` functions are listed in Tables `r knitr::asis_output(ifelse(knitr::is_latex_output(), '\\ref{tab:tbl-num-la}','\\@ref(tab:tbl-num)'))` and `r knitr::asis_output(ifelse(knitr::is_latex_output(), '\\ref{tab:tbl-cat-la}','\\@ref(tab:tbl-cat)'))` respectively. The `data`, `id`, `values`, and `time` arguments are required for both functions. The `ncat`, `breaks`, `group_scaling`, and `scaling` arguments are for the user to customize the data scaling calculation in the `anim_prep()` function. In the `anim_prep_cat` function, the users can customize the order of the category to be shown on the y-axis using the `order` argument. Then, for further customization regarding how the final visualization looks. The `group`, `label`, and `time_dependent` arguments can be adjusted.

@@

The `cat_change` data in the \CRANpkg{animbook} package is used to illustrate how to use this package. It has five categories, A to E, over two time points. Since the data is already in the categorized data structure, the `anim_prep_cat()` function is used. The output from this function is a class called `categorized`.

@@

Figure \@ref(fig:top-many) shows the flexibility to use various options, such as `subset` or `relation`, to customize the plot appearance. Here `subset = "top"` is used to specify category A for the final time point, and `relation = "many_one"` is used to specify that all categories flowing into A at time point 2 are shown.
